Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-25 numerical-conduction.tex
Jeff Oishi's changes.
Edited section 8 (conduction, star formation + feedback routines).

\enzo\ implements the equations of isotropic heat conduction in a manner
similar to that of \citet{2007ApJ...664..135P}.  
The isotropic flux of heat is given by:
\begin{equation}
\myvec{F} = -\kappa_{\rm sp} \grad T
\end{equation}
where $\kappa_{\rm sp}$ is the Spitzer conduction coefficient, $4.6 \times 10^{-7}$~T$^{5/2}$ erg
s$^{-1}$~cm$^{-1}$~K$^{-1}$ \citep{1962pfig.book.....S}, here we are using a
value for the Coulomb logarithm, $\log \Lambda = 37.8$, that is
appropriate for the intracluster medium \citep{1988xrec.book.....S}.
It is quite possible that the local heat flux computed in this way 
can become unphysically large in
the high-temperature, low-density cluster regime when using this
formulation; therefore, we take into account the saturation of the heat
flux \citep{1977ApJ...211..135C} at a maximum level of
\begin{equation}
F_{sat} \simeq 0.4 n_e k_b T \left( \frac{2 k_b T}{\pi m_e} \right)^{1/2}.
\end{equation}
To ensure a smooth transition between
the Spitzer and saturated regimes, we define an effective
conductivity using the formalism of \citet{1988xrec.book.....S}
\begin{equation}
\kappa_{eff} = \frac{\kappa_{\rm sp}}{1 + 4.2 \lambda_e / \ell_T} \; ,
\end{equation}
where $\lambda_e$ is the electron mean free path and $\ell_T \equiv T
/ |\grad T|$ is the characteristic length scale of the local
temperature gradient.  We also assume that the conductivity of the
plasma can be described in terms of an effective conductivity, which
can be expressed as a fraction f$_{\rm sp}$ of the Spitzer conductivity
(where f$_{\rm sp} \leq 1.0$ are considered physical values).  This takes
into account physical processes below the resolution limit of the
simulation (such as tangled magnetic fields) that can suppress heat
transport.

Thermal conduction in a plasma can be strongly affected by the
presence of magnetic field lines, which may suppress heat
flow perpendicular to the magnetic field.  In that case, we allow for
heat transport only parallel to the magnetic field lines in
magnetohydrodynamic simulations.  Mathematically, this is treated as:
\begin{equation}
\myvec{F} = -\kappa_{\rm sp} \hat{b} (\hat{b} \cdot \grad T)
\end{equation}
where $\hat{b}$ is the unit vector denoting the direction of the
magnetic field.  As with the isotropic thermal conduction, we allow a
multiplicative factor f$_{\rm sp}$ to take into account the possible
behavior of magnetic fields below the resolution limit of the
simulation.

>>>>

\enzo\ implements the equations of isotropic heat conduction in a manner
similar to that of \citet{2007ApJ...664..135P}.  
The isotropic flux of heat is given by
\begin{equation}
\myvec{F} = -\kappa_{\rm sp} \grad T
\end{equation}
where $\kappa_{\rm sp}$ is the Spitzer conduction coefficient, $4.6 \times 10^{-7}$~T$^{5/2}$ erg
s$^{-1}$~cm$^{-1}$~K$^{-1}$ \citep{1962pfig.book.....S}.  In this
situation we are using a
value for the Coulomb logarithm, $\log \Lambda = 37.8$, that is
appropriate for the intracluster medium \citep{1988xrec.book.....S} --
in astrophysically-relevant, fully ionized plasmas this value varies
by no more than 50\% (see, e.g., Smith et al. 2013, submitted).
It is quite possible that the local heat flux computed in this way 
can become unphysically large in
the high-temperature, low-density cluster regime when using this
formulation; therefore, we take into account the saturation of the heat
flux \citep{1977ApJ...211..135C} at a maximum level of
\begin{equation}
F_{sat} \simeq 0.4 n_e k_b T \left( \frac{2 k_b T}{\pi m_e} \right)^{1/2}.
\end{equation}
To ensure a smooth transition between
the Spitzer and saturated regimes, we define an effective
conductivity using the formalism of \citet{1988xrec.book.....S}
\begin{equation}
\kappa_{eff} = \frac{\kappa_{\rm sp}}{1 + 4.2 \lambda_e / \ell_T} \; ,
\end{equation}
where $\lambda_e$ is the electron mean free path and $\ell_T \equiv T
/ |\grad T|$ is the characteristic length scale of the local
temperature gradient.  We also assume that the conductivity of the
plasma can be described in terms of an effective conductivity, which
can be expressed as a fraction f$_{\rm sp}$ of the Spitzer conductivity
(where f$_{\rm sp} \leq 1.0$ are considered physically realistic values).  This takes
into account physical processes below the resolution limit of the
simulation, such as tangled magnetic fields, that can suppress heat
transport.

Thermal conduction in a plasma can be strongly affected by the
presence of magnetic field lines, which may suppress heat
flow perpendicular to the magnetic field.  In that case, we allow for
heat transport only parallel to the magnetic field lines in
magnetohydrodynamic simulations.  Mathematically, this is treated as:
\begin{equation}
\myvec{F} = -\kappa_{\rm sp} \myvec{b} (\myvec{b} \cdot \grad T)
\end{equation}
where $\myvec{b}$ is the unit vector denoting the direction of the
magnetic field.  As with the isotropic thermal conduction, we allow a
multiplicative factor f$_{\rm sp}$ to take into account the possible
behavior of magnetic fields below the resolution limit of the
simulation.

### 2013-06-25 numerical-starformation.tex
Edited section 8 (conduction, star formation + feedback routines).

Due to the computationally unfeasible number of stars in a galaxy
($10^{11}$) and the lack of detailed understanding of star formation, a
number of phenomenological star formation models are included in
\enzo.
%The \enzo\ code includes approximately ten models for star formation
%and feedback.  
Broadly speaking, these methods all work in similar
ways: At a specified time interval, all grid cells that are at the
highest local level of refinement (i.e., that have no child cells) are
examined to see if they meet a set of criteria for star formation.
This may simply be a baryon density threshold, but may include more
complex tests, such as an examination of local cooling and dynamical
time scales, molecular hydrogen fraction, metallicity, and converging
gas flows.  If the star formation criteria are met, some mass of gas
is taken away from the cell in question and a ``star particle'' with
the same mass is placed in the center of that cell with the same
velocity as the removed gas.  This star particle is then
allowed to inject mass, momentum, thermal energy, metals, and possibly
magnetic fields and/or cosmic ray populations into its local
environment.  In general, the particle is treated as an ensemble of
stars, with feedback properties occuring over time according to the
assumed initial mass function of the stellar population.

@@

During this timestep, the star particle returns
metal-enriched gas and thermal energy from supernovae and from stellar
winds.  Since massive stars have very short lifetimes, we assume that
there is an immediate feedback of some fraction $f_{\rm SN}$ of the rest
energy from the stars into the baryon gas, such
that $E_{\rm add} = f_{\rm SN} \Delta m_{\rm sf} c^2$, where c is the speed of
light.  In addition, a fraction $f_{Z*}$ of the stellar mass is fed back 
in the form of metals.  Finally, a fraction of the mass $f_{m*}$
is added back into the gas
along with momentum in order to simulate the mass ejection from
all stars (not just supernovae).

There are six user-defined parameters in this algorithm: three deal
with the star formation ($\eta$, $m_{\rm *min}$ and $t_{\rm
  dyn,min}$), and three deal with feedback ($f_{\rm SN}$, $f_{Z*}$ and
$f_{m*}$).  Some of these parameters are completely free, while others
can be guided by observation or theory.  For example, the supernova
feedback parameter, $f_{\rm SN}$, can be constrained assuming that, for
every $200 M_\odot$ of stars created, one supernova occurs, and this
event feeds back approximately $10^{51}$ ergs of thermal energy,
giving:

@@

The methods described in the previous two sections are generally used
in simulations that have very poor resolution, $\Delta x \ga 1$~kpc.
At this physical scale, individual star forming regions are not
resolved, so all uncertainty about the behavior of molecular clouds is
folded into a density threshold for star formation.  In calculations
with much higher resolution, however (on the order of a few pc), individual
molecular clouds can be resolved, thus rendering these approximations
invalid.  To that end, \citet{2012ApJ...749...36K} have implemented
a star formation algorithm that is specifically geared to
high-resolution cosmological simulations of galaxy formation, where
the formation of molecular hydrogen is followed directly, and stars
are allowed to form at the highest level of refinement when the local
H$_2$ fraction exceeds a pre-determined threshold.  Cells are examined
every root grid timestep, $\Delta t_0$, and cells at the highest level
that exceed the H$_2$ limit form star particles with masses
proportional to the inferred mass of molecular hydrogen in the
star-forming region
\citep{2008ApJ...689..865K,2009ApJ...693..216K,2010ApJ...709..308M}.
The mass of the particle is calculated as:
\begin{equation} 
m_p = \epsilon \rho_{\rm gas} (\Delta x_{m})^3 \frac{\Delta t_0}{t_*}
\end{equation}
where $\Delta x_{m}$ is the resolution of the maximum level of
refinement, $\epsilon$ is an efficiency parameter with a standard
value of 0.01 \citep[as motivated by][]{2007ApJ...654..304K}, and
t$_*$ is the local free-fall timescale.

>>>>

Due to the computationally unfeasible number of stars in a galaxy
($10^{11}$) and the lack of detailed understanding of star formation, a
number of phenomenological star formation models are included in
\enzo.
%The \enzo\ code includes approximately ten models for star formation
%and feedback.  
Broadly speaking, these methods all work in similar
ways: At a specified time interval, all grid cells that are at the
highest local level of refinement (i.e., that have no child cells) are
examined to see if they meet a set of criteria for star formation.
This may simply be a baryon density threshold, but can also include more
complex tests, such as an examination of local cooling and dynamical
time scales, molecular hydrogen fraction, metallicity, and converging
gas flows.  If the star formation criteria are met, some mass of gas
is taken away from the cell in question and a ``star particle'' with
the same mass is placed in the center of that cell with the same
velocity as the removed gas.  This star particle is then
allowed to inject mass, momentum, thermal energy, metals, and possibly
magnetic fields and/or cosmic ray populations into its local
environment.  In general, the particle is treated as an ensemble of
stars, with feedback properties occuring over time according to the
assumed initial mass function of the stellar population.

@@

During this timestep, the star particle returns
metal-enriched gas and thermal energy from supernovae and from stellar
winds.  Since massive stars have very short lifetimes, we assume that
there is an immediate feedback of some fraction $f_{\rm SN}$ of the rest
energy from the stars into the gas, such
that $E_{\rm add} = f_{\rm SN} \Delta m_{\rm sf} c^2$, where c is the speed of
light.  In addition, a fraction $f_{Z*}$ of the stellar mass is fed back 
in the form of metals.  Finally, a fraction of the mass $f_{m*}$
is added back into the gas
along with momentum in order to simulate the mass ejection from
all stars (not just supernovae).

There are six user-defined parameters in this algorithm: three deal
with star formation ($\eta$, $m_{\rm *min}$ and $t_{\rm
  dyn,min}$), and three deal with feedback ($f_{\rm SN}$, $f_{Z*}$ and
$f_{m*}$).  Some of these parameters are completely free, while others
can be guided by observation or theory.  For example, the supernova
feedback parameter, $f_{\rm SN}$, can be constrained by assuming that, for
every $200 M_\odot$ of stars created, one supernova occurs, and this
event feeds back approximately $10^{51}$ ergs of thermal energy,
giving:

@@

The methods described in the previous two sections are generally used
in simulations that have very poor resolution, $\Delta x \ga 1$~kpc.
At this physical scale, individual star forming regions are not
resolved, so all uncertainty about the behavior of molecular clouds is
folded into a density threshold for star formation.  In calculations
with much higher resolution, however (on the order of a few pc), individual
molecular clouds can be resolved, thus rendering these approximations
invalid.  To that end, \citet{2012ApJ...749...36K} have implemented
a star formation algorithm that is specifically geared to
high-resolution cosmological simulations of galaxy formation, where
the formation of molecular hydrogen is followed directly and stars
are allowed to form at the highest level of refinement when the local
H$_2$ fraction exceeds a pre-determined threshold.  Cells are examined
every root grid timestep, $\Delta t_0$, and cells at the highest level
that exceed the H$_2$ limit form star particles with masses
proportional to the inferred mass of molecular hydrogen in the
star-forming region
\citep{2008ApJ...689..865K,2009ApJ...693..216K,2010ApJ...709..308M}.
The mass of the particle is calculated as:
\begin{equation} 
m_p = \epsilon \rho_{\rm gas} (\Delta x_{m})^3 \frac{\Delta t_0}{t_*}
\end{equation}
where $\Delta x_{m}$ is the resolution of the maximum level of
refinement, $\epsilon$ is an efficiency parameter with a standard
value of 0.01 \citep[as motivated by][]{2007ApJ...654..304K}, and
t$_*$ is the local free-fall timescale.
